---
title: "Create-mounds-training-data"
author: "Jakob Mørup"
date: "2025-05-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Installing and loading packages
```{r}
#install.packages("pacman")
pacman::p_load(terra, tidyverse, sf, measurements, stringr, raster,purrr)


```

loading the KAZ_moundattributes and VisibleID to pick out the visible mounds to create bounding boxes for training the U-Net model later.
```{r}
Mounds <- read.csv("../data/KAZ_moundattributes.csv") #Read in mound attributes

visibleID <- read.delim("../data/visibleID.txt") 
colnames(visibleID) <- "TRAP_ID"   #Read in list of visible mounds and change column name to match the one in mounds


visibleMounds <- inner_join(Mounds,visibleID) %>% mutate(
                                                  max_dim = pmax(Length, Width),
                                                  size_bucket = case_when(
                                                  max_dim <= 10 ~ "small",
                                                  max_dim <= 40 ~ "medium",
                                                  TRUE ~ "large"
                                                  )
                                                  )  #inner join to only pick the ones in the visible list, then find the maximum dimension for each, and bucket                                                        them according to that.


```


for each mounds I now make a bounding box of different dimensions for each size bucket. 
this is to make sure the background doesn't completely drown the signal in noise for the small and medium mounds.
these will be used to fine tune different heads on a U-Net segmentation model (together with patches of similar sizes without mounds in them)



generate bounding boxes.
```{r}
#####################################
# SETTING THE BUFFER SIZES, THESE CONTROL THE SIZE OF THE BOUNDING BOXES FOR SMALL/MEDUIM/LARGE MOUNDS
#####################################
buffer_sizes <- c(6,24,78) #these are set at 20% bigger than the threshold and for large mounds 20% bigger than 130 which is the max dimension of the largest mound
#this makes the mounds (assuming they are circular) fill at max ~71% of the area and the minimum would be ~21% for a mound that just makes the large category.
#the idea is that having some background around the mound is good for training the segmentation model, but if the mound is only a tiny part of the image, it will learn
#background noise instead. 
#thresholds are x<=10,x<=40,40<=x, where x is the largest dimension of the mound, for small/medium and large respectively.
```
Trying a modified version of Adelas approach

```{r}
mounds2 <- st_read("../data/KAZ_mounds.shp")
# filter the noticeable ones (2+ meters)
mounds2 <- mounds2 %>% 
  left_join(Mounds, by = c("TRAP_Code"="TRAP_ID"))


visibleMounds2 <- inner_join(mounds2,visibleID, by = c("TRAP_Code"="TRAP_ID")) %>% mutate(
                                                  max_dim = pmax(Length, Width),
                                                  size_bucket = case_when(
                                                  max_dim <= 10 ~ "small",
                                                  max_dim <= 40 ~ "medium",
                                                  TRUE ~ "large"
                                                  )
                                                  )  #inner join to only pick the ones in the visible list, then find the maximum dimension for each, and bucket                                                        them according to that.

#add buffer sizes as col
visibleMounds2 <- visibleMounds2 %>%
  mutate(buffer = case_when(
    size_bucket == "small" ~ buffer_sizes[1],
    size_bucket == "medium" ~ buffer_sizes[2],
    size_bucket == "large" ~ buffer_sizes[3]
  ))


#pulling out center coordinates for creating label masks, together with max_dim
center_coords <- st_coordinates(visibleMounds2)
colnames(center_coords) <- c("center_x","center_y")
visibleMounds2 <- cbind(visibleMounds2, center_coords)

```
```{r}
morevisibleMounds2 <- mounds2 %>% mutate(
                                                  max_dim = pmax(Length, Width),
                                                  size_bucket = case_when(
                                                  max_dim <= 10 ~ "small",
                                                  max_dim <= 40 ~ "medium",
                                                  TRUE ~ "large"
                                                  )
                                                  )  #inner join to only pick the ones in the visible list, then find the maximum dimension for each, and bucket                                                        them according to that.

#add buffer sizes as col
morevisibleMounds2 <- morevisibleMounds2 %>%
  mutate(buffer = case_when(
    size_bucket == "small" ~ buffer_sizes[1],
    size_bucket == "medium" ~ buffer_sizes[2],
    size_bucket == "large" ~ buffer_sizes[3]
  ))


#pulling out center coordinates for creating label masks, together with max_dim
center_coords <- st_coordinates(morevisibleMounds2)
colnames(center_coords) <- c("center_x","center_y")
morevisibleMounds2 <- cbind(morevisibleMounds2, center_coords)

```

```{r}

#create empty list to hold bounding box polygons
polylist <- list()
polylist2 <- list()


#function to take in a point and create a square polygon based on buffer size, centered on this point
make_square_bbox <- function(pt, size) {
  coords <- st_coordinates(pt)
  x <- coords[1]
  y <- coords[2]
  box <- matrix(c(
    x - size, y - size,
    x + size, y - size,
    x + size, y + size,
    x - size, y + size,
    x - size, y - size
  ), ncol = 2, byrow = TRUE)
  return(st_polygon(list(box))) 
}


#looping over each point and adding the resulting polygon from above function to the list
for (i in 1:248) {
  a <- visibleMounds2[i,]$geometry
  b <- visibleMounds2[i,]$buffer
  polylist[i] <- make_square_bbox(a,b)
}

for (i in 1:773) {
  a <- morevisibleMounds2[i,]$geometry
  b <- morevisibleMounds2[i,]$buffer
  polylist2[i] <- make_square_bbox(a,b)
}

#turning the list of matrices into a Simple Feature Column (sfc) 
sfc_polygons <- st_sfc(lapply(polylist, function(mat) st_polygon(list(mat))))
sfc_polygons2 <- st_sfc(lapply(polylist2, function(mat) st_polygon(list(mat))))


#binding this new sfc to my dataset
visibleMounds3 <- cbind(visibleMounds2, sfc_polygons)
morevisibleMounds3 <- cbind(morevisibleMounds2, sfc_polygons2)


# Convert it back to an sf object in case something broke along the way
visibleMounds3 <- st_as_sf(visibleMounds3)
morevisibleMounds3 <- st_as_sf(morevisibleMounds3)

# Explicitly set the active geometry column to the new polygon column
st_geometry(visibleMounds3) <- "geometry.1"
st_geometry(morevisibleMounds3) <- "geometry.1"

# Check if everything is correctly structured
print(st_geometry_type(visibleMounds3))
print(st_geometry_type(morevisibleMounds3))

#remove the POINT geometry, and rename the POLYGON geometry to "geometry"
visibleMounds4 <- visibleMounds3 %>% 
  dplyr::select(-geometry) %>% 
  rename(geometry = geometry.1) %>% 
  st_as_sf()

morevisibleMounds4 <- morevisibleMounds3 %>% 
  dplyr::select(-geometry) %>% 
  rename(geometry = geometry.1) %>% 
  st_as_sf()

#apply the right crs
st_crs(visibleMounds4) <- st_crs(visibleMounds2)
st_crs(morevisibleMounds4) <- st_crs(morevisibleMounds2)


#write to a file
st_write(visibleMounds4, "bounding_boxes.geojson", delete_dsn = TRUE)
st_write(morevisibleMounds4, "more_bounding_boxes.geojson", delete_dsn = TRUE)


```

